\Title

### 2023-02-09 Grimaldi-etal-BiolCybernetics.tex
Methods : 2.3

%In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Please refer to suggested Data Availability Statements in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}. If the study did not report any data, you might add ``Not applicable'' here. Chen
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NOTATIONS
% Pixel world
\newcommand{\presynaddr}{a} % pre address
\newcommand{\postsynaddr}{b} % post address
\newcommand{\numevent}{N_{ev}} % total number of events
\newcommand{\presynaddrspace}{\mathcal{A}} %presynaptic address space
\newcommand{\postsynaddrspace}{\mathcal{B}} %postsynaptic address space
\newcommand{\Npol}{N_\text{p}} % number of polarity
\newcommand{\Nneuron}{N_\text{n}} % number of output neurons in the layer
\newcommand{\arank}{r} % address index
\newcommand{\bias}{\beta} % bias for the MLR model
\newcommand{\synapse}{\mathcal{S}} % synapse
\newcommand{\synapticweight}{w} % synaptic weight
\newcommand{\synapticdelay}{\delta} % synaptic delay
\newcommand{\ranksyn}{s} % synapse index
\newcommand{\Nsyn}{N_{s}} % total number of synapses
\newcommand{\activeweights}{\mathcal{W}} 
\newcommand{\timev}{t} % time
\newcommand{\polev}{p} % polarity
\newcommand{\event}{\epsilon} % event
\newcommand{\eventstream}{\xi} % stream of events
\newcommand{\TS}{S} % time surface
\newcommand{\neuron}{\mathbf{n}} % neuron in the SNN (defined by the spatial position and the channel)
\newcommand{\postneuron}{\mathbf{m}} % post synaptic neuron in the SNN (defined by the spatial position and the kernel)
\newcommand{\channel}{\mathbf{p}} % channel
\newcommand{\layer}{\mathbf{L}} % layer
\newcommand{\ms}{\si{\milli\second}}%
\newcommand{\us}{\si{\micro\second}}%
\newcommand{\timecontext}{T} % time context (cf HOTS) matrice gathering last event times
\newcommand{\current}{I} % post synaptic current
\newcommand{\volt}{u} % membrane potential
\newcommand{\volts}{V} % matrix of membrane potentials
\newcommand{\gain}{\gamma} % homeostatic gain
\newcommand{\simil}{\beta} % similarity value
\newcommand{\Nclass}{N_\text{class}} % number of classes for MLR:
\newcommand{\Nx}{N_\text{X}}
\newcommand{\Ny}{N_\text{Y}}
\newcommand{\Ntime}{N_\text{T}}
\newcommand{\kernel}{K} % convolution kernel
%\newcommand{\kernelind}{\mathbf{k}} % indice of the kernel
\newcommand{\kernelind}{k} % indice of the kernel
\newcommand{\Kx}{K_\text{x}}
\newcommand{\Ky}{K_\text{y}}
\newcommand{\Ktime}{K_\text{t}}
\newcommand{\classiflayer}{\mathbf{C}}
\newcommand{\class}{c} % class k of the MLR
\newcommand{\lrweights}{\theta} % matrix of MLR weights
\newcommand{\lrtrue}{y} % true value of the prediction for MLR
\newcommand{\loss}{J} % cost function for MLR
\newcommand{\softmax}{\sigma}
\newcommand{\actfreq}{f}
\newcommand{\decision}{\hat{y}}
\newcommand{\colorsec}{black}
\newcommand{\colorsubsec}{black}
\newcommand{\speed}{v}
\newcommand{\Nspeed}{N_v}
% Example definitions.
% --------------------
\def\x{{\mathbf x}}
\def\L{{\cal L}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Task definition: fast motion detection}\label{sec:task}

@@

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf - Motion Detection Task.} {\bf (Left)}~We use large natural images ($512\times512$) in which an aperture ($128\times128$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, the view axis moves according to a stepwise random walk. We show an example path of the trajectory for $200$ time steps (green line). {\bf (Right)}~Snapshot of the synthetic event stream at different time steps. The dynamics of the sub-image as a function of time produces a naturalistic movie. This can be transformed into an event-based representation. Mimicking the retina, this representation encodes pixel-per-pixel proportional increases or decreases in luminance in the image, i.e. ON (red) and OFF (blue) events. In the lower left corner of the images, the translation vector is shown in red as one of the possible classes of motion. Note the change in the direction of motion between the second and third image.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. % For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. 
For this purpose, we selected a database of 600 natural images that were previously used to study the statistics of natural images~\citep{perrinet_edge_2015}. Note that these have been pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band. This process is known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $256 \times 256$ in size, and we will extract sub-images of size $128 \times 128$ positioned around the center of gaze at each time step. We will discretize time in $1~\ms$ bins and produce movies of duration $N_T = 250~\ms$. To avoid boundary effects, we will randomly position the full trajectory in image space such that the subimage is translated using the position given by the trajectory at each time step without touching the borders. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a subpixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity to correspond to a movement of one pixel per frame (i.e., per time bin).

@@

Let's formalize the heterogeneous delays spiking neural network (HD-SNN). Each neuron $\postsynaddr \in \postsynaddrspace$ connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses. Each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. Scanning all neurons $\postsynaddr$, we thus define the full set of $\Nsyn$ synapses, 
as  $\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$, where each synapse is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$, a weight $\synapticweight_\ranksyn$, and a delay $\synapticdelay_\ranksyn$. This defines the full connectivity of the HD-SNN model. Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]} $, or the emitting field of a presynaptic neuron $\synapse_\presynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]}$. As a consequence, a postsynaptic neuron $\postsynaddr$ receives an event stream which is multiplexed by the synapses of its receptive field. It results in a weighted event stream: 
$\event_\postsynaddr = \{(\presynaddr_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}$. Unlike the binary stream of events from the signal as input, the postsynaptic neurons receives a weighted stream of event such that for each event $\arank$ its value is $\synapticweight_\arank = \synapticweight_\ranksyn$ when $\presynaddr_\arank = \presynaddr_\ranksyn$. This new stream of events is by nature ordered in time as events reach the soma of post-synaptic neurons. In particular, when post-synaptic neurons are activated on their soma by this spatio-temporal motif, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner. This model is defined in this subsection in all generality and the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions. In the next subsection we describe an implementation of our model adapted for the motion detection task.

\note{I can make a small figure with event streams of different colors for the input one, the multiplexed one etc and the response in the bottom}

@@

From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input as the boolean matrix $A \in \{0, 1 \}^{N\times T}$, where $N$ is the number of neurons in $\presynaddrspace$ and $T$ is the number of time bins. In this simplified model, we will consider that heterogeneous delays are limited in range such that the synaptic set can be represented as a kernel by the dense matrix $\kernel^\postsynaddr$ of size $(\Ntime, \Nx, \Ny)$, where $\Ntime$ is the number of delays and $\Nx$ and $\Ny$ are the number of pixels in both spatial dimensions. For each neuron $\postsynaddr$, $\kernel^\postsynaddr$ gives the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in \synapse^\postsynaddr}, \kernel^\postsynaddr(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) = \synapticweight_\ranksyn$. Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. Using this dense representation, the integration of the spike train as input by the synapses of neuron $\postsynaddr$ is:

\begin{equation}\label{eq:kernel_b}
\forall \presynaddr \in \presynaddrspace^{\synapse^\postsynaddr}, \forall \timev \in T, \; \;
\mathcal{C}^\postsynaddr(\presynaddr,\timev)
= \sum_{\ranksyn  \in [1, N_{\synapse^\postsynaddr}]} \kernel^\postsynaddr(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn)
\end{equation}
where $\presynaddrspace^{\synapse^\postsynaddr}$ is the address space of the receptive field of neuron $\postsynaddr$ and $N_{\synapse^\postsynaddr}$ is the number of synapses belonging to neuron $\postsynaddr$. To benefit from invariance to location observed in CNN and to simplify the implementation of the model, one assumes that kernels should be similar across different positions and define a convolution operator. The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$. The computational brick used in equation~\eqref{eq:kernel_b} can be extended to every position, in space and time, of the boolean matrix $A$:

\begin{equation}\label{eq:conv}
\forall \presynaddr \in \presynaddrspace, \forall \timev \in T, \; \;
\mathcal{C}_\class(\presynaddr,\timev)
= \sum_{\ranksyn  \in [1, N_{\synapse^\postsynaddr}]} \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn)
\end{equation}

This shows that $\mathcal{C}_\class$ is a spatiotemporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (see Figure~\ref{fig:model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream.
%
The same type of spatiotemporal filtering is used as a preprocessing stage of a pattern recognition algorithm~\cite{ghosh_spatiotemporal_2019}. And in \cite{sekikawa_constant_2018}, an efficient 3D convolutional algorithm succeeds a motion estimation task. By assuming a locally constant velocity the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity. Here, we keep the analogy with spiking neurons and we try to use as few heuristics as possible to observe the emergence of spiking motifs in the spatiotemporal kernels. If so, this precise spatiotemporal patterns prove to be of interest for neural computations and one can assume that biological neurons make use of this information as well.

\note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}

The 3D convolution represents the linear integration of the spike train as input. The activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. 
It transforms this list of weights into a probability with the following formula:

\begin{equation}\label{eq:mlr}
\forall \presynaddr \in \presynaddrspace, \forall \timev \in T, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \; \presynaddr, \timev) =
\frac 1 Z
{\exp  (\mathcal{C}_\class(\presynaddr, \timev) +\bias_\class) }
\end{equation}

where $\bias_\class$ is the bias linked to class $\class$ and $Z$ is the partition function. 
In our MLR model, a probability value, for each class (i.e. each motion direction), is predicted for each event at address $\presynaddr_\arank$ and at time $\timev_\arank$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class, keeping the same timing as the input event.

@@

\note{describe the heuristics inspired by neuroscience}

\note{describe the parameters of the model}

\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/HD-SNN.png}
    \textbf{- HD-SNN illustration. From left to right:} A 3D representation of the event stream as output of the DVS. A 3D convolution is applied, on the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. There are as many kernels as the possible number of motion classes in the recording. For clarity, only $4$ kernels are illustrated and some weights are not represented in the voxel grid. For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the non-linearity of the MLR model (i.e. softmax function). The output of the MLR gives a probability for each class, associated to a specific kernel, to emit an output event, or spike. With an appropriate spiking mechanism, here a winner-takes-all, we obtain, as output of the HD-SNN model, a new spike train with the different spikes associated to a specific motion class. }
    \label{fig:model}
\end{figure}

>>>>

%In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Please refer to suggested Data Availability Statements in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}. If the study did not report any data, you might add ``Not applicable'' here. Chen
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% NOTATIONS
% Pixel world
\newcommand{\presynaddr}{a} % pre address
\newcommand{\postsynaddr}{b} % post address
\newcommand{\numevent}{N_{ev}} % total number of events
\newcommand{\presynaddrspace}{\mathcal{A}} %presynaptic address space
\newcommand{\postsynaddrspace}{\mathcal{B}} %postsynaptic address space
\newcommand{\Npol}{N_\text{p}} % number of polarity
\newcommand{\Nneuron}{N_\text{n}} % number of output neurons in the layer
\newcommand{\arank}{r} % address index
\newcommand{\bias}{\beta} % bias for the MLR model
\newcommand{\synapse}{\mathcal{S}} % synapse
\newcommand{\synapticweight}{w} % synaptic weight
\newcommand{\synapticdelay}{\delta} % synaptic delay
\newcommand{\ranksyn}{s} % synapse index
\newcommand{\Nsyn}{N_{s}} % total number of synapses
\newcommand{\activeweights}{\mathcal{W}} 
\newcommand{\timev}{t} % time
\newcommand{\polev}{p} % polarity
\newcommand{\event}{\epsilon} % event
\newcommand{\eventstream}{\xi} % stream of events
\newcommand{\TS}{S} % time surface
\newcommand{\neuron}{\mathbf{n}} % neuron in the SNN (defined by the spatial position and the channel)
\newcommand{\postneuron}{\mathbf{m}} % post synaptic neuron in the SNN (defined by the spatial position and the kernel)
\newcommand{\channel}{\mathbf{p}} % channel
\newcommand{\layer}{\mathbf{L}} % layer
\newcommand{\ms}{\si{\milli\second}}%
\newcommand{\us}{\si{\micro\second}}%
\newcommand{\timecontext}{T} % time context (cf HOTS) matrice gathering last event times
\newcommand{\current}{I} % post synaptic current
\newcommand{\volt}{u} % membrane potential
\newcommand{\volts}{V} % matrix of membrane potentials
\newcommand{\gain}{\gamma} % homeostatic gain
\newcommand{\simil}{\beta} % similarity value
\newcommand{\Nclass}{N_\text{class}} % number of classes for MLR:
\newcommand{\Nx}{N_\text{X}}
\newcommand{\Ny}{N_\text{Y}}
\newcommand{\Ntime}{N_\text{T}}
\newcommand{\kernel}{\mathcal{K}} % convolution kernel
%\newcommand{\kernelind}{\mathbf{k}} % indice of the kernel
\newcommand{\kernelind}{k} % indice of the kernel
\newcommand{\Kx}{K_\text{X}}
\newcommand{\Ky}{K_\text{Y}}
\newcommand{\Ktime}{K_\text{T}}
\newcommand{\classiflayer}{\mathbf{C}}
\newcommand{\class}{c} % class k of the MLR
\newcommand{\lrweights}{\theta} % matrix of MLR weights
\newcommand{\lrtrue}{y} % true value of the prediction for MLR
\newcommand{\loss}{J} % cost function for MLR
\newcommand{\softmax}{\sigma}
\newcommand{\actfreq}{f}
\newcommand{\decision}{\hat{y}}
\newcommand{\colorsec}{black}
\newcommand{\colorsubsec}{black}
\newcommand{\speed}{v}
\newcommand{\Nspeed}{N_v}
% Example definitions.
% --------------------
\def\x{{\mathbf x}}
\def\L{{\cal L}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Task definition: motion detection in a synthetic naturalistic event stream}\label{sec:task}

@@

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf - Motion Detection Task.} {\bf (Left)}~We use large natural images ($512\times512$) in which an aperture ($128\times128$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, the view axis moves according to a stepwise random walk. We show an example path of the trajectory for $200$ time steps (green line). {\bf (Right)}~Snapshot of the synthetic event stream at different time steps. The dynamics of the sub-image as a function of time produces a naturalistic movie. This can be transformed into an event-based representation. Mimicking the retina, this representation encodes pixel-per-pixel proportional increases or decreases in luminance in the image, i.e. ON (red) and OFF (blue) events. In the lower left corner of the images, the translation vector is shown in red as one of the possible classes of motion. Note the change in the direction of motion between the second and third image.}
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. % For this purpose, we selected a database of 10 natural grayscale images that are commonly used to study the statistics of natural images~\citep{olshausen_emergence_1996}. 
For this purpose, we selected a database of 600 natural images that were previously used to study the statistics of natural images~\citep{perrinet_edge_2015}. Note that these have been pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band. This process is known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $256 \times 256$ in size, and we will extract sub-images of size $64 \times 64$ positioned around the center of gaze at each time step. We will discretize time in $1~\ms$ bins and produce movies of duration $N_T = 128~\ms$. To avoid boundary effects, we will randomly position the full trajectory in image space such that the subimage is translated using the position given by the trajectory at each time step without touching the borders. The translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a sub-pixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the velocity to correspond to a movement of one pixel per frame (i.e., per time bin).

@@

Let's formalize the heterogeneous delays spiking neural network (HD-SNN). Each neuron $\postsynaddr \in \postsynaddrspace$ connects to presynaptic afferent from $\presynaddrspace$. In biology, a single cortical neuron has generally several thousands of synapses. Each synapse may be defined by its synaptic weight and its delay, that is, the time it takes for one spike to travel from the presynaptic neuron's soma to that of the postsynaptic neuron. A postsynaptic neuron $\postsynaddr \in \postsynaddrspace$ is then described by the synaptic weights connecting it to a presynaptic afferent from $\presynaddrspace$ but also by the set of possible delays. Note that a neuron may contact an afferent neuron with multiple different delays. Scanning all neurons $\postsynaddr$, we thus define the full set of $\Nsyn$ synapses, 
as  $\synapse = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn)\}_{\ranksyn \in [1,\Nsyn]}$, where each synapse is associated to a presynaptic address $\presynaddr_\ranksyn$, a postsynaptic address $\postsynaddr_\ranksyn$, a weight $\synapticweight_\ranksyn$, and a delay $\synapticdelay_\ranksyn$. This defines the full connectivity of the HD-SNN model.

Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]} $, or the emitting field of a presynaptic neuron $\synapse_\presynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]}$. As a consequence, a postsynaptic neuron $\postsynaddr$ receives an event stream which is multiplexed by the synapses of its receptive field. It results in a weighted event stream: 
$\event_\postsynaddr = \{(\presynaddr_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}$. Unlike the binary stream of events from the signal as input, the postsynaptic neurons receives a weighted stream of event such that for each event $\arank$ its value is $\synapticweight_\arank = \synapticweight_\ranksyn$ when $\presynaddr_\arank = \presynaddr_\ranksyn$. This new stream of events is by nature ordered in time as events reach the soma of post-synaptic neurons. In particular, when post-synaptic neurons are activated on their soma by this spatio-temporal motif, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner. This model is defined in this subsection in all generality and the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions. In the next subsection we describe an implementation of our model adapted for the motion detection task.

% \note{I can make a small figure with event streams of different colors for the input one, the multiplexed one etc and the response in the bottom}

@@

From the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input as the boolean matrix $A \in \{0, 1 \}^{\Nx \times \Ny \times \Ntime}$, defined in space coordinates $x$ and $y$ and time $t$. To benefit from invariance to location observed in CNN and to simplify the implementation of the model, one assumes that kernels should be similar across different positions and define a convolution operator. % The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$. 
Here, we will consider that heterogeneous delays are limited in range such that the synaptic set can be represented as a kernel for each class $\class$ of the supervision task by the dense matrix $\kernel_\class$ of size $(\Kx, \Ky, \Ktime)$, where $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions and  $\Ktime$ is the number of delays. To keep an analogy with the HD-SNN model, for each neuron $\postsynaddr$ of position $(x, y)$ and class $\class$, $\kernel_\class$ gives the weights as a function of presynaptic address and delay: $\forall {\ranksyn \in \synapse^\postsynaddr}, \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) = \synapticweight_\ranksyn$. Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. Using this dense representation, the processing of $A$ by the model can be written as:
\begin{equation}\label{eq:kernel_b}
\forall x, y, \forall \timev \in T, \; \;
\mathcal{C}_\class(x, y, \timev)
= \sum_{\delta_x, \delta_y, \synapticdelay} \kernel_\class(\delta_x, \delta_y, \synapticdelay) \cdot A(x - \delta_x, y - \delta_y, \timev-\synapticdelay)
\end{equation}
% The computational brick used in equation~\eqref{eq:kernel_b} can be extended to every position, in space and time, of the boolean matrix $A$:
% \begin{equation}\label{eq:conv}
% \forall \presynaddr \in \presynaddrspace, \forall \timev \in T, \; \;
% \mathcal{C}_\class(\presynaddr,\timev)
% = \sum_{\ranksyn  \in [1, N_{\synapse^\postsynaddr}]} \kernel_\class(\presynaddr_\ranksyn, \synapticdelay_\ranksyn) \cdot A(\presynaddr, \timev-\synapticdelay_\ranksyn)
% \end{equation}
This shows that $\mathcal{C}_\class$ is a spatiotemporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (see Figure~\ref{fig:model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time.
This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream.

The same type of spatiotemporal filtering is used as a preprocessing stage of a pattern recognition algorithm~\cite{ghosh_spatiotemporal_2019}. And in \cite{sekikawa_constant_2018}, an efficient 3D convolutional algorithm succeeds a motion estimation task. By assuming a locally constant velocity, the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity. Here, we keep the analogy with spiking neurons, and we try to use as few heuristics as possible to observe the emergence of spiking motifs in the spatiotemporal kernels. If so, this precise spatiotemporal patterns prove to be of interest for neural computations and one can assume that biological neurons make use of this information as well. 
% \note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}
The 3D convolution represents the linear integration of the spike train as input. The activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Take-All network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value, for each class (i.e. each motion direction), is predicted for each event at position $x, y$ and at time $\timev$ as a softmax function of the linear combination of the list of events on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. Using the kernels, it transforms the input raster plot into a probability with the following formula:
\begin{equation}\label{eq:mlr}
    \forall x, y, \forall \timev \in T, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \;  x, y, \timev) =
\frac {\exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class) }{Z(x, y, \timev)}
\end{equation} 
where $\bias_\class$ is the bias linked to class $\class$ and $Z(x, y, \timev)=\sum_{\class \in [1, N_\class]} \exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class)$ is the partition function. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class.

@@

% \note{describe the heuristics inspired by neuroscience}

% \note{describe the parameters of the model}

\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=\linewidth]{figures/HD-SNN.png}
    \caption{
    \textbf{- HD-SNN illustration. From left to right:} A 3D representation of the event stream as output of the DVS. A 3D convolution is applied, on the dense representation of the input, along the $X$ and $Y$ spatial axes (defining the DVS pixel grid) and the temporal axis. There are as many kernels as the possible number of motion classes in the recording. For clarity, only $4$ kernels are illustrated and some weights are not represented in the voxel grid. For each voxel on the dense representation of the event stream as input, the output of the 3D convolution is processed by the non-linearity of the MLR model (i.e. softmax function). The output of the MLR gives a probability for each class, associated to a specific kernel, to emit an output event, or spike. With an appropriate spiking mechanism, here a winner-takes-all, we obtain, as output of the HD-SNN model, a new spike train with the different spikes associated to a specific motion class. }
    \label{fig:model}
\end{figure}

\subsection{Supervised learning of the motion detection task}
